Parsing Postfix log files

### 2007-08-13 doc/logparser.tex
First round of corrections after proof reading; there are more to come.

    Parsing Postfix logs is much more difficult than it first appears but
    it is possible to achieve a high degree of accuracy in understanding
    the logs and reconstructing the actions taken by Postfix to generate
    the logs.  This paper describes the process required, documenting the
    parsing algorithm and rules, explaining the difficulties encountered,
    with reference to an implementation which stores data gathered from the
    logs in an SQL database.  The gathered data can then be used to
    optimise current anti-spam measures, provide a baseline to test new
    anti-spam measures against, or to produce statistics showing how
    effective those measures are.

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication arises because the list of rejection messages is not fixed:
new messages can be added by the administrator with custom checks; every
Real-time Black List returns a different explanatory message; policy
servers may have different messages depending on the characteristics of the
connection; there are many ways in which the log lines may differ between
servers.  This paper documents the difficult process of parsing Postfix
logs, presenting a program which parses logs and places the resulting data
into a database.  The gathered data can then be used to optimise current
anti-spam measures, provide a baseline to test new anti-spam measures
against, or to produce statistics showing how effective those measures are.
There are numerous other uses for such data: improving server performance
by identifying troublesome destinations and reconfiguring appropriately or
dedicating transports to those destinations; identifying regular high
volume uses (e.g.\ customer newsletters) and restricting those uses to
off-peak times; detecting virus outbreaks which propagate via email; as a
base for billing customers on a shared server.

@@

Section~\ref{rules} discusses the parsing rules in detail, explaining their
structure, giving an example rule and sample data it matches successfully
against, followed by a discussion of rule efficiency concerns.

@@

Section~\ref{limitations-improvements} lists the limitations of the program
and/or the algorithm, then suggests some ways of dealing with the
limitations, with the goal of improving parsing and reproducing the journey
a mail takes through the internals of Postfix.

@@

Section~\ref{bibliography} is the bibliography, containing the resources
used in developing the algorithm, writing the program and preparing this
paper.  Also listed are some additional resources expected to be helpful in
understanding SMTP, Postfix, anti-spam techniques, or the paper.

Section~\ref{graphs} contains graphs illustrating the topics discussed in
section~\ref{rule efficiency}.

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are worth using in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.\footnote{Section~\ref{introduction} lists some more example
uses.}

@@

\begin{verbatim}
SELECT name, description, restriction_name, rule_order
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY rule_order DESC;
\end{verbatim}

@@

\begin{verbatim}
SELECT name, description, restriction_name, rule_order,
        (rule_order * 100.0 /
            (SELECT SUM(rule_order)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY rule_order DESC;
\end{verbatim}

@@

\begin{verbatim}
SELECT name, description, restriction_name, rule_order,
        (rule_order * 100.0 /
            (SELECT SUM(rule_order)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
        AND rule_order < 100
    ORDER BY rule_order ASC;
\end{verbatim}

@@

The database populated by this program provides a simple interface to
Postfix logs.  Although the interface is a database schema, it is in effect
quite similar to the API provided by a library: it insulates both user and
provider of the API from changes in the implementation of the other.  The
algorithm implemented by the program can be improved; support can be added
for earlier or later releases of Postfix; bugs can be fixed or limitations
removed from the program.  Statistics and/or graphs can be generated from
the database; new restrictions can be tested and the results inspected;
trends in the fight against spam can emerge from historical data saved in
the database.  Using a database simplifies writing programs which need to
interact with the data in several ways:

@@

    \item Libraries exist for the majority of programming languages
        allowing access to databases, whereas if this parser was written
        as a library a new interface layer would need to be written for
        every programming language using the API\@.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of variable complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

@@

            \item Other tables can be added to the database.

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and cause the correct action to be executed; they also have several
fields which aid the user in understanding what each rule parses.  The
rules are described in detail in section~\ref{rules} and the fields in
particular in section~\ref{rule attributes}, and that discussion won't be
repeated here.  The data saved from parsing the logs is divided into two
tables: connections and results.

@@

    \item [helo] The hostname used in the HELO command.  The HELO
        occasionally changes during a connection, presumably because spam
        or virus senders think it's a good idea.  By default Postfix only
        logs this when it rejects an SMTP command, but it is trivial to log
        this for every connection:

        \texttt{smtpd\_helo\_restrictions = \newline
        \hspace*{2em} check\_helo\_access regexp:/etc/postfix/log\_helo.regex}

        Where \texttt{/etc/postfix/log\_helo.regex} contains:

        \texttt{/./ WARN Logging HELO for diagnostic purposes}

    \item [queueid] The queueid of the mail, or \textit{NOQUEUE\/} if a
        mail wasn't accepted.

@@

Every log line corresponding to Postfix performing an action other than
INFO will have an entry in the results table, e.g.\ rejecting an SMTP
command, delivering a mail, or bouncing a mail.  Any log line where the
Postfix action is INFO is not interesting in and of itself, but provides
additional information which will be saved with other results.

@@

    \item [connection\_id] A reference to the entry in the connections
        table this results entry belongs to.

@@

    \item [warning] Postfix can log a warning instead of rejecting an SMTP
        command --- a facility that is quite useful for testing new
        restrictions.  This field will be 1 if the log line parsed was a
        warning rather than a real rejection, or 0 for a real rejection or
        any other result.

    \item [smtp\_code] The SMTP code extracted from the log line.  In
        general an SMTP code is only present for a rejection or final
        delivery; results missing an SMTP code will duplicate the SMTP code
        of other results.  Some final delivery log lines don't contain an
        SMTP code: in those cases the code is faked based on the success or
        failure represented by the log line.

@@

The Simple Mail Transfer Protocol, originally defined in RFC
821~\cite{RFC821} and later in RFC 2821~\cite{RFC2821}, is used for
transferring mail between Mail Transport Agents (MTAs).  It is a simple,
human readable, plain text protocol, making it quite easy to test and debug
problems with it.  Despite the simplicity many virus and/or spam sending
programs fail to implement it properly, so requiring strict adherence to
the protocol specification is beneficial in protecting against spam and
viruses.\footnote{This violates the principle of \textit{Be liberal in what
you accept, and conservative in what you send\/} from RFC
760~\cite{rfc760}, but unfortunately that principle was written in a
friendlier time.}  A typical SMTP conversation resembles the following (the
lines starting with a three digit number are sent by the server, all other
lines are sent by the client):

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of severity and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained shortly.} as the ultimate
in flexibility.  Administrators can also supply their own rejection
messages to make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the journey an email takes through
Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
pattern: 90\% of the time the journey is simple, but the other 10\% of the
time requires 90\% of the code.\footnote{These numbers don't have a solid
scientific basis, they're based on gut feeling from writing and debugging
the software.}

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table.} will check
whether the IP address of the connected client matches the left hand side
of each line in the file and will return the right hand side of the first
matching line.  Other restrictions determine their result in different
ways, e.g.\ \texttt{reject\_rbl\_client some.rbl.com} checks whether the IP
address of the client is present in the listed RBL\@.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue with the next restriction in the list, allowing more specific
cases to be handled differently from more general cases.  When the result
is another restriction Postfix will evaluate the new restriction, allowing
restrictions to be chosen based on the client IP address, HELO hostname,
sending address, recipient address, etc.  The administrator can define new
restrictions as a list of existing restrictions, allowing arbitrarily long
and complex sequences of lookups and restrictions.

@@

    \item Greylisting~\cite{greylisting}, is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on the second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        retrying after a temporary failure is uneconomical for a spammer,
        but that a legitimate mail server must retry.  Sadly spammers are
        using increasingly complex and well written programs to distribute
        spam, frequently using an ISP provided SMTP server from a
        compromised machine on the ISP's network.  Greylisting is slowly
        becoming less useful, but it does block a large percentage of spam
        mail at the moment.

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser: rules identify the line and return some data,
which the algorithm (the YACC part) receives and uses when performing the
requested action.  Rules are solely concerned with identifying a line and
extracting data from it, whereas the algorithm's task is to follow the
journey each mail takes through Postfix, piecing the data together into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.  Readers familiar with context-free grammars may find
section~\ref{comparison against context-free grammars} helpful.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of Log Mail Analyzer is
quite different from his parser: it attempts to present correlated log data
in a form suitable for a systems administrator to search using grep(1),
producing a CSV file\footnote{Comma-Separated Value, a very simple format
where a record is stored in a single line, with fields separated by a comma
or other punctuation symbol.  Problems with CSV files include the
difficulty in escaping separators, providing multiple values for a field
(e.g.\ multiple recipients), adding new fields, and lack of a bundled schema
describing the fields.} which can subsequently be imported into a MySQL or
Berkeley DB database.  The data stored is limited to time and date,
hostname and IP address of client, sender and recipient addresses, and the
SMTP code.  The decision to support both CSV and Berkeley DB appears to
have been a limiting factor: the values which can be stored are limited,
it's difficult to link different records (foreign keys in SQL enable this
easily), and ad-hoc queries require extensive code to be written.  Berkeley
DB is not an SQL database, it only supports (key, value) pairs; in the main
table the value used is a CSV line, and the key is an integer referred to
by other tables.  The secondary by-sender, by-recipient, by-date and by-IP
tables use the sender/recipient/date/IP as the key, and a CSV list of
integers referring to the main table is used as the value.  In effect this
re-implements SQL foreign keys, but without the full functionality offered
by even the most basic of SQL databases.  It also requires custom code to
search on some combination of the above.

@@

@@

    \item [description] Something must have occurred to cause Postfix to
        log each line (e.g.\ a remote client connecting causes a connection
        line to be logged).  The description field is generally used to
        describe the action causing the log lines this rule matches.

@@

    \item [program] The program (postfix/smtp, postfix/smtpd, postfix/qmgr,
        etc.) whose log lines the rule applies to.  This avoids needlessly
        trying rules which won't match the line, or worse, might match
        unintentionally.  Rules whose program is \texttt{*} will be tried
        against any lines which aren't parsed by program specific rules.

    \item [regex] The regular expression\footnote{See
        http://en.wikipedia.org/wiki/Regular\_expression for more
        information about regular expressions in general.} to match the log
        line against.  The regex will first have several keywords expanded:
        this simplifies reading and writing rules, avoids needless
        repetition of complex regex components, allows the components
        to be corrected and/or improved in one location, and makes each
        regex largely self-documenting.

@@

        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
        \newline server\_hostname and helo.\newline} \texttt{result\_cols:
        sender, recipient, smtp\_code, data \newline and child.}

@@

    \item [hits] is an efficiency measure.  This counter is maintained for
        every rule and incremented each time the rule successfully matches.
        At the start of each run the program sorts the rules in descending
        hits, and at the end of the run updates every rule's hits.
        Assuming that the distribution of log lines is reasonably
        consistent between log files, rules matching more commonly
        occurring log lines will be tried before rules matching less
        commonly occurring log lines, lowering the program's execution
        time.  Rule ordering for efficiency is discussed in
        section~\ref{rule ordering for efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: rules
        will be tried in order of priority, overriding hits.  This allows
        more specific rules to take precedence over more general rules.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's regex, generate an NFA, and compare all NFAs
to determine if any overlap.  The author has not attempted to do this: such
a project would probably qualify for a PhD, and may involve solving the
Halting Problem and circumventing the Church-Turing Thesis.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line determines which rule
matches.  Overlapping rules are frequently a requirement, allowing a more
specific rule to match some lines and a more general rule to match the
remainder, e.g.\ separating smtp delivery to a filtering proxy from smtp
delivery to the rest of the world.  The algorithm provides a facility for
handling overlapping rules: the priority field in each rule (defaults to
zero).  Rules are tried in order of priority, highest first, overriding the
hits field maintained by the parser.  Negative priorities may also be
useful for catchall rules.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See section~\ref{rule efficiency} for more
        details of sorting the rules.}  Parse a number of log files using
        normal sorting, dump a textual representation of the connections
        and results tables, and delete everything from those tables.
        Repeat with shuffled and reversed sorting.  If there are no
        overlapping rules the tables from each run will be identical;
        differences indicate overlapping rules, and the rule\_ids which
        differ determine which rules overlap.  Unfortunately this method
        cannot prove the absence of overlapping rules; it can detect
        overlapping rules --- but only if there are log lines which match
        more than one rule.

@@

\begin{verbatim}
NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 
  550 <foo@bar.baz>: Sender address rejected: Domain not found;
  from=<foo@bar.baz> to=<info@example.com> proto=SMTP
  helo=<smtp.bar.baz>
\end{verbatim}

@@

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the line the regex will be matched against the
        line; if the match is successful the action will be called, if not
        the next rule will be tried.

@@

    \item [queueid:] is used the extract the queueid matched by the regex,
        so that the mail can be found by queueid and actions performed on
        it.

@@

Additional fields are captured by \_\_RESTRICTION\_START\_\_, hence the
fields in result\_cols and connection\_cols start at 5.

@@

It's probably apparent by now that tracking re-injected mails is the single
most complex part of the algorithm.  Section~\ref{complications} has most
of the previous discussion, should you wish to review it.

@@

>>>>

    Parsing Postfix logs is much more difficult than it first appears but
    it is possible to achieve a high degree of accuracy in understanding
    the logs, and thus reconstructing the actions taken by Postfix to
    generate the logs.  This paper describes the creation of a parser,
    documenting the parsing algorithm and rules, explaining the
    difficulties encountered and the solutions developed, with reference to
    an implementation which stores data gleaned from the logs in an SQL
    database.  The gathered data can then be used to optimise current
    anti-spam measures, provide a baseline to test new anti-spam measures
    against, or to produce statistics showing how effective those measures
    are.

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or users.\footnote{This was the
author's first real foray into processing Postfix logs.}  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix logs, where each log line
was added on an ad hoc basis as a requirement was discovered or new
functionality was added.  Further complication arises because the set of
rejection messages is not fixed: new messages can be added by the
administrator with custom checks; every Real-time Black List returns a
different explanatory message; policy servers may log different messages
depending on the characteristics of the connection; there are many ways in
which the log lines may differ between servers.  This paper documents the
difficult process of parsing Postfix logs, presenting a program which
parses logs and places the resulting data into a database.  The gathered
data can then be used to optimise current anti-spam measures, provide a
baseline to test new anti-spam measures against, or to produce statistics
showing how effective those measures are.  There are numerous other uses
for such data: improving server performance by identifying troublesome
destinations and reconfiguring appropriately or dedicating transports to
those destinations; identifying regular high volume uses (e.g.\ customer
newsletters) and restricting those uses to off-peak times; detecting virus
outbreaks which propagate via email; as a base for billing customers on a
shared server.  Preserving the raw data enables users to develop a
multitude of uses far beyond those conceived of by the author.

@@

Section~\ref{rules} discusses the parsing rules in detail, explaining their
structure, using an example rule and sample data it matches successfully
against, followed by a discussion of rule efficiency concerns.

@@

Section~\ref{limitations-improvements} lists the limitations of the program
and/or the algorithm, then suggests some ways of dealing with the
limitations, with the goal of improving parsing and reproduction of the
journey a mail takes through Postfix.

@@

The bibliography contains references to the resources used in developing
the algorithm, writing the program and preparing this paper.  Also listed
are some additional resources expected to be helpful in understanding SMTP,
Postfix, anti-spam techniques, or the paper.

Section~\ref{graphs} contains graphs illustrating the topics discussed in
section~\ref{rule efficiency}, rule efficiency.

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are beneficial in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.\footnote{Section~\ref{introduction} lists some more example
uses.}

@@

\begin{verbatim}
SELECT name, description, restriction_name, hits_total
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY hits_total DESC;
\end{verbatim}

@@

\begin{verbatim}
SELECT name, description, restriction_name, hits_total,
        (hits_total * 100.0 /
            (SELECT SUM(hits_total)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY hits_total DESC;
\end{verbatim}

@@

\begin{verbatim}
SELECT name, description, restriction_name, hits,
        (hits * 100.0 /
            (SELECT SUM(hits)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
        AND hits < 100
    ORDER BY hits ASC;
\end{verbatim}

@@

The database populated by this program provides a simple interface to
Postfix logs.  Although the interface is a database schema, it is in effect
quite similar to the API provided by a library: it insulates both user and
provider of the API from changes in the implementation of the API\@.  The
algorithm implemented by the parser can be improved; support can be added
for earlier or later releases of Postfix; bugs can be fixed or limitations
removed from the parser.  Statistics and/or graphs can be generated from
the database; new restrictions can be tested and the results inspected;
trends in the fight against spam can emerge from historical data saved in
the database.  Using a database simplifies writing programs which need to
interact with the data in several ways:

@@

    \item Libraries exist for the majority of programming languages
        allowing access to databases, whereas if this parser was written as
        a shared library a new interface layer would need to be written for
        every programming language using the API\@.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of varying complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

@@

            \item Other tables can be added to the database, e.g.\ to cache
                historical data.

@@

XXX SHOULD THIS SECTION BE MOVED ELSEWHERE\@?

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in section~\ref{rules} and the fields in particular in
section~\ref{rule attributes}, and that discussion won't be repeated here.
The data saved from parsing the logs is divided into two tables described
below: connections and results.

@@

    \item [helo] The hostname used in the HELO command.  The HELO
        occasionally changes during a connection, presumably because spam
        or virus senders think it's a good idea.  By default Postfix only
        logs the HELO when it rejects an SMTP command, but it is quite easy
        to rectify this:

\label{logging helo}

            \item Modify \texttt{smtpd\_helo\_restrictions} in
                \texttt{/etc/postfix/main.cf} to contain\newline
                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}

        The purpose of logging the HELO in
        \texttt{smtpd\_helo\_restrictions} is to log only once per mail; it
        also seems like the natural place.

    \item [queueid] The queueid of the mail if the conection represents an
        accepted mail, or \textit{NOQUEUE\/} otherwise.

@@

Every log line corresponding to Postfix performing an action other than
INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
SMTP command, delivering a mail, or bouncing a mail.  Any log line where
the Postfix action is INFO is not interesting in and of itself, but
provides additional information which will be saved with other results;
IGNORED log lines don't even provide useful information.

@@

    \item [connection\_id] A reference to the entry in the connections
        table this results table entry belongs to.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an SMTP command --- a
        facility that is quite useful for testing new restrictions.  This
        field will be 1 if the log line parsed was a warning rather than a
        real rejection, or 0 for a real rejection or any other result.

    \item [smtp\_code] The SMTP code associated with the log line.  In
        general an SMTP code is only present for a rejection or final
        delivery; results missing an SMTP code will duplicate the SMTP code
        of other results in the connection.  Some final delivery log lines
        don't contain an SMTP code: in those cases the code is faked based
        on the success or failure represented by the log line.

@@

The Simple Mail Transfer Protocol, originally defined in RFC
821~\cite{RFC821} and updated in RFC 2821~\cite{RFC2821}, is used for
transferring mail between Mail Transport Agents (MTAs).  It is a simple,
human readable, plain text protocol, making it quite easy to test and debug
problems with it.  Despite the simplicity many virus and/or spam sending
programs fail to implement it properly, so requiring strict adherence to
the protocol specification is beneficial in protecting against spam and
viruses.\footnote{This violates the principle of \textit{Be liberal in what
you accept, and conservative in what you send\/} from RFC
760~\cite{rfc760}, but unfortunately that principle was written in a
friendlier time.}  A typical SMTP conversation resembles the following (the
lines starting with a three digit number are sent by the server, all other
lines are sent by the client):

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive and extensible optional
anti-spam restrictions, allowing an administrator to deploy those
restrictions which they judge suitable for their site's needs, rather than
a fixed set chosen by Postfix's author.  These restrictions can be
selectively applied, combined and bypassed on a per-client, per-recipient
or per-sender basis, allowing varying levels of severity and/or
permissiveness.  Postfix leverages simple lookup tables to support
arbitrarily complicated user-defined sequences of restrictions and
exceptions, with policy servers\footnote{Policy servers will be explained
shortly.} as the ultimate in flexibility.  Administrators can also supply
their own rejection messages to make it clear to senders why exactly their
mail was rejected.  Unfortunately this flexibility has a cost: complexity
in the logs generated.  While it is easy to use \texttt{grep(1)} to
determine the fate of an individual email, following the journey an email
takes through Postfix can be quite difficult.  The logs tend to follow a
90\%-10\% pattern: 90\% of the time the journey is simple, but the other
10\% of the time requires 90\% of the code.\footnote{These numbers don't
have a solid scientific basis, they're based on gut feeling from writing
and debugging the software.}

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table.} will check
whether the IP address of the connected client matches the left hand side
of each line in the file and will return the right hand side of the first
matching line.  Other restrictions determine their result in different
ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks whether the
IP address of the client is present in the listed RBL\@.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue with the next restriction in the list, allowing more specific
cases to be handled differently from more general cases.  When the result
is another restriction Postfix will evaluate the new restriction, allowing
restrictions to be chosen based on the client IP address, HELO hostname,
sender address, recipient address, etc.  The administrator can define new
restrictions as a list of existing restrictions, allowing arbitrarily long
and complex sequences of lookups and restrictions.

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on the second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        retrying after a temporary failure is uneconomical for a spammer,
        but that a legitimate mail server must retry.  Sadly spammers are
        using increasingly complex and well written programs to distribute
        spam, frequently using an ISP provided SMTP server from a
        compromised machine on the ISP's network.  Greylisting is slowly
        becoming less useful, but it does block a large percentage of spam
        mail at the moment.

@@

XXX REWRITE THIS

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser: rules identify the line and return some data,
which the algorithm (the YACC part) receives and uses when performing the
requested action.  Rules are solely concerned with identifying a line and
extracting data from it, whereas the algorithm's task is to follow the
journey each mail takes through Postfix, piecing the data together into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of Log Mail Analyzer is
quite different from this parser: it attempts to present correlated log
data in a form suitable for a systems administrator to search using
grep(1), producing a CSV file\footnote{Comma-Separated Value, a very simple
format where a record is stored in a single line, with fields separated by
a comma or other punctuation symbol.  Problems with CSV files include the
difficulty in escaping separators, providing multiple values for a field
(e.g.\ multiple recipients), adding new fields, and lack of a bundled
schema describing the fields.} which can subsequently be imported into a
MySQL or Berkeley DB database.  The data stored is limited to time and
date, hostname and IP address of client, sender and recipient addresses,
and the SMTP code.  The decision to support both CSV and Berkeley DB
appears to have been a limiting factor: the values which can be stored are
limited, it's difficult to link different records (foreign keys in SQL
enable this easily), and ad-hoc queries require extensive code to be
written.  Berkeley DB is not an SQL database, it only supports (key, value)
pairs; in the main table the value used is a CSV line, and the key is an
integer referred to by other tables.  The secondary by-sender,
by-recipient, by-date and by-IP tables use the sender/recipient/date/IP as
the key, and a CSV list of integers referring to the main table is used as
the value.  In effect this re-implements SQL foreign keys, but without the
full functionality offered by even the most basic of SQL databases.  It
also requires custom code to search on some combination of the above.

@@

XXX EXPLAIN WHY ASO IS MORE COMPREHENSIVE, WHAT IT DOES BETTER AND WHY\@.
MAKE THIS A COMPARISON INSTEAD OF A CRITIQUE\@.

@@

    \item [description] Something must have occurred to cause Postfix to
        log each line (e.g.\ a remote client connecting causes a connection
        line to be logged).  This field describes the action causing the
        log lines this rule matches.

@@

    \item [program] The program (\texttt{postfix/smtpd},
        \texttt{postfix/qmgr}, etc.) whose log lines the rule applies to.
        This avoids needlessly trying rules which won't match the line, or
        worse, might match unintentionally.  Rules whose program is
        \texttt{*} will be tried against any lines which aren't parsed by
        program specific rules.

    \item [regex] The regular expression\cite{wikipedia-regex, perlre} to
        match the log line against.  The regex will first have several
        keywords expanded: this simplifies reading and writing rules;
        avoids needless repetition of complex regex components; allows the
        components to be corrected and/or improved in one location; and
        makes each regex largely self-documenting.

@@

        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
        \newline \hspace*{2em} server\_hostname and helo.\newline}
        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
        \newline \hspace*{2em} pid\_regex, data and child.}

@@

    \item [hits] is an efficiency measure.  This counter is maintained for
        every rule and incremented each time the rule successfully matches.
        At the start of each run the program sorts the rules in descending
        order of hits, and at the end of the run updates every rule's hits.
        Assuming that the distribution of log lines is reasonably
        consistent between log files, rules matching more commonly
        occurring log lines will be tried before rules matching less
        commonly occurring log lines, lowering the program's execution
        time.  Rule ordering for efficiency is discussed in
        section~\ref{rule ordering for efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: rules
        will be tried in order of priority, overriding hits.  This allows
        more specific rules to take precedence over more general rules
        (described in section~\ref{overlapping rules}).

@@

\label{overlapping rules}

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's regex, generate an NFA, and compare all NFAs
to determine if any overlap.  The author has not attempted to do this: such
a project would probably qualify for a PhD, and may involve solving the
Halting Problem and circumventing the Church-Turing Thesis.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line determines which rule
matches.  Overlapping rules are frequently a requirement, allowing a more
specific rule to match some lines and a more general rule to match the
majority, e.g.\ separating smtp delivery to a filtering proxy from smtp
delivery to the rest of the world.  The algorithm provides a facility for
ordering overlapping rules: the priority field in each rule (defaults to
zero).  Rules are tried in order of priority, highest first, overriding the
hits field maintained by the parser.  Negative priorities may also be
useful for catchall rules.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See section~\ref{rule efficiency} for more
        details of sorting the rules.}  Parse a number of log files using
        normal sorting, dump a textual representation of the connections
        and results tables, and delete everything from those tables.
        Repeat with shuffled and reversed sorting.  If there are no
        overlapping rules the tables from each run will be identical;
        differences indicate overlapping rules, and the rule\_ids which
        differ determine which rules overlap.  Unfortunately this method
        cannot prove the absence of overlapping rules; it can detect
        overlapping rules --- but only if there are log lines in the input
        files which match more than one rule.

@@

\begin{verbatim}
NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 
  550 <foo@example.com>: Sender address rejected:
  Domain not found; from=<foo@example.com>
  to=<info@example.net> proto=SMTP
  helo=<smtp.example.com>
\end{verbatim}

@@

\vspace{1em}

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the line the regex will be matched against the
        line; if the match is successful the action will be executed, if not
        the next rule will be tried.  If the program-specific rules don't
        match the log line, the generic rules will be tried in the same
        way.

@@

    \item [queueid:] is used the extract the queueid matched by the regex,
        so that the correct mail can be found by queueid and actions
        performed on it.

@@

Additional fields are captured by \_\_RESTRICTION\_START\_\_, hence the
fields in result\_cols and connection\_cols start at 5 in the example.

@@

@@

The actions:
